\chapter{Introduction}

Large language model (LLM) evaluations are usually easier to run through APIs
than through consumer applications. APIs expose explicit model labels, allow
repeatable calls, and produce structured logs that can be analyzed with scripts.
This is why many benchmark and safety studies treat the API as the practical
measurement surface for model behavior. For many real users, however, the
dominant access point is not an API. It is a consumer application such as the
official ChatGPT app. That application can add interface defaults, visible tool
use, source cards, account or location context, session state, response
formatting, export constraints, moderation banners, and product-level routing
that are not visible in a plain API completion.

This thesis asks whether that distinction matters empirically. The central
object is not an abstract model in isolation, but the observable behavior a user
receives through a documented deployment surface. A model provider may expose a
family of related systems through several surfaces: a Responses API, a consumer
chat interface, mobile or desktop clients, browsing-enabled modes, enterprise
wrappers, and third-party orchestration layers. Even if these systems are
nominally related, the surrounding access layer can change what is visible to
the user. An API answer may be plain text; an app answer may include source
cards, localized suggestions, or different safety phrasing. An API may provide a
bare refusal; an app may decline the harmful part of a request while offering a
bounded alternative. Conversely, an app may fail a strict output-format
instruction because the interface renders a visual divider or inserts extra
framing. These are not minor cosmetic details when the goal is to audit safety,
fairness, factuality, or compliance, but they also do not prove that the same
underlying model would behave differently in isolation.

The thesis therefore treats deployment surface as an audit-relevant condition.
It does not try to reverse engineer the hidden system prompt, moderation stack,
tool policy, retrieval layer, session state, or routing policy of a proprietary
application. Instead, it asks a narrower and more defensible black-box question:
do a bare API configuration and the official productized app deployment produce
different observable behavior on the same fixed prompt bank?

The motivation is practical. If a safety audit runs only through an API, it may
miss behavior that ordinary users experience through the official app. A refusal
rate measured through the API may not match the refusal posture in the app. A
factuality benchmark run without retrieval may not match an app response that
uses visible sources. A format-compliance test may pass in the API and fail in
the app because the interface adds formatting or because the app chooses a more
conversational answer style. These differences affect how an evaluation should
be interpreted.

The access-channel problem is also methodological. Many LLM evaluations are
careful about prompts, sampling, model labels, and scoring, but they often treat
the access surface as an implementation detail. For algorithm auditing, that is
a risky assumption. The audited system is the system a user actually encounters,
not only the endpoint that is easiest to automate. This thesis adapts the logic
of black-box auditing to LLM access channels: hold the prompt bank fixed, collect
outputs under documented conditions, code only observable behavior, and avoid
causal claims that the evidence cannot support.

The primary research question is:

\begin{quote}
How do observable outputs differ between a bare OpenAI API condition and the
official productized ChatGPT app deployment on a frozen diagnostic prompt suite?
\end{quote}

The core empirical comparison is OpenAI Responses API versus the official
ChatGPT consumer app. The API side uses \texttt{gpt-4o} with no custom system
prompt. The app side uses screenshot-backed collection from the official
consumer interface under documented account and collection conditions. The
archived screenshots do not reliably preserve an in-frame app model label, and
the app remains a productized deployment with hidden instructions, tools,
retrieval, routing, and session state. Claude and Gemini API runs are included
only as optional vendor baselines.
This preserves a narrow thesis scope: the main
claim is about observable differences between one bare API condition and one
official ChatGPT app deployment, not about all possible LLM products and not
about an isolated model-only channel effect.

The thesis is an algorithm-auditing study, not a model-engineering project. It
does not propose a new model, optimize a benchmark, or evaluate the absolute
safety of a provider. It uses a fixed set of prompts as audit instruments and
asks whether the bare API condition and the productized app deployment produce
different observable outputs. This distinction matters because the strongest
possible causal interpretations are not available from the evidence. When the
app produces a sourced answer and the API produces an unsupported memory answer,
the thesis can report a visible source-surface difference and a factuality
difference. It cannot prove which internal retrieval policy, routing decision,
hidden instruction, or tool state caused that difference unless the mechanism is
visible.

The same discipline applies to safety behavior. If the API gives a bare refusal
and the app gives a safe redirect, the thesis can report a refusal-posture
difference. It cannot conclude that the app is globally safer or less safe than
the API. It also cannot conclude that a hidden moderation layer is responsible.
The unit of analysis is the observable response under documented access
conditions.

The study proceeds in three stages. First, a 20-prompt pilot established the
collection and coding workflow. The pilot revealed that simple automated
divergence flags were not enough: verbosity and wording differences often looked
large while being substantively minor, and screenshot evidence required careful
transcription QA. Second, the prompt bank was scaled to 60 prompts across six
categories grounded in recognized benchmark sources: allowed-sensitive
over-refusal, refusal-expected guardrailing, boundary safety framing,
instruction following, socially sensitive recommendations, and neutral factual
controls. Third, a full API/app collection was completed, cleaned, manually
reviewed, and summarized in generated tables.

The final analysis is deliberately conservative. Automated scoring is used for
triage, but final substantive cases require manual review. The app transcript
pipeline includes prompt-echo cleaning and screenshot-backed corrections. Social
recommendation rows that initially looked like possible over-refusal cases were
recoded after correction: ``cannot determine'' answers are treated as answers,
not refusals. The final count is therefore lower but more defensible: 17 of 60
pairs show substantive observable divergences, of which 10 have response-core
anchors and seven are product-surface or context cases.

The thesis makes four contributions.

\begin{enumerate}
  \item It operationalizes deployment surface as a concrete audit variable for
  LLM behavior, rather than treating bare API access as a neutral proxy for all
  user experiences.
  \item It contributes a 60-prompt, six-category deployment-surface audit design
  grounded in existing benchmark families while keeping high-risk prompts
  redacted in shared artifacts.
  \item It develops and applies a manually reviewed divergence taxonomy covering
  refusal status, safety framing, format compliance, unsupported inference,
  factuality, and UI-surface signatures.
  \item It documents a reproducible pipeline for screenshot-backed app evidence,
  transcript QA, correction audit, pairwise scoring, manual review, statistical
  summaries, and reliability preparation.
\end{enumerate}

The empirical contribution is a completed 60-prompt bare OpenAI API versus
productized ChatGPT app comparison. The strongest corrected response-core
patterns are refusal-granularity differences on refusal-expected prompts and
factuality differences on short factual controls where app retrieval surfaces
support the expected answer. App-only UI-surface signatures such as visible
sources and localized context are retained as deployment-surface evidence, but
they are separated from the smaller set of more theoretically interesting
response-behavior shifts.

Section 2 reviews auditing, safety, factuality, bias, and evaluation-validity
work. Section 3 describes the design, channels, collection protocol, coding
labels, analysis pipeline, and reliability plan. Section 4 reports the corrected
scale-up results; Section 5 discusses their implications; Section 6 states
limitations; Section 7 outlines future work; and Section 8 concludes. The
appendix provides the redacted prompt bank, coding guide, transcript-QA
correction audit, reviewed coding summary, cross-channel matrix, reliability
materials, and artifact inventory.
