\section{Related Work}

This thesis sits at the intersection of algorithm auditing and LLM evaluation.
It draws on audit methodology for studying opaque deployed systems, safety and
refusal benchmarks for constructing risky and benign prompts, instruction and
factuality benchmarks for verifiable controls, and bias benchmarks for socially
sensitive inference tasks. The review is organized around the methodological
problem that motivates the thesis: existing benchmarks are often carefully
designed, but their results are usually tied to a single access surface, most
often an API. The sections below first situate the thesis in the auditing and
accountability tradition, then review the LLM systems and benchmark families the
prompt bank draws on, and finally narrow to the access-channel question itself.

\subsection{Algorithm Auditing and Accountability}

Algorithm auditing treats a deployed computational system as an object of
empirical study. Rather than relying only on documentation or provider claims,
auditors send controlled inputs to a system and compare outputs across cases,
conditions, or identities. Sandvig et al.\ \cite{sandvig_auditing} frame this
style of research as a way to detect discrimination and systematic behavior in
opaque internet platforms, adapting the logic of classical housing-discrimination
audits to software systems. Their central insight is methodological: when the
internal mechanism is unavailable, one can still construct controlled tests that
reveal patterns in observable outputs. Diakopoulos \cite{diakopoulos_accountability}
develops the same idea as a journalistic practice, using systematic probing to
hold ``computational power structures'' accountable, and Kroll et al.\
\cite{kroll_accountable} analyze, from a legal and technical perspective, why
post-hoc external scrutiny is necessary when systems cannot be made fully
transparent in advance.

Empirical audits in this tradition have repeatedly shown that deployed
commercial systems behave differently from what documentation or marketing
suggests. Buolamwini and Gebru's Gender Shades study \cite{gender_shades} found
large intersectional accuracy disparities in commercial gender-classification
products, and the follow-up by Raji and Buolamwini \cite{raji_actionable} showed
that publicly naming audited products measurably changed vendor behavior.
Obermeyer et al.\ \cite{obermeyer_bias} dissected a widely used healthcare risk
algorithm and located its racial bias in a proxy-variable choice, a finding that
was only visible by comparing observable predictions against outcomes. These
studies matter for the present thesis not because of their domains but because
of their epistemic position: each treated the deployed product, not the
underlying model in isolation, as the unit of audit.

A second strand of this literature concerns how audits should be embedded in
practice and what they can validly claim. Raji et al.\ \cite{raji_gap} propose
an end-to-end internal auditing framework intended to close the gap between
high-level AI principles and concrete deployment decisions. Selbst et al.\
\cite{selbst_abstraction} warn that fairness work fails when it abstracts away
the sociotechnical context in which a system runs, and Suresh and Guttag
\cite{suresh_guttag} catalogue how distinct harms enter at distinct stages of
the machine-learning life cycle, including deployment. Binns
\cite{binns_fairness} connects fairness measurement to longer-standing debates
in political philosophy, and Barocas and Selbst \cite{barocas_selbst} analyze
why facially neutral data practices can produce disparate impact. The textbook
treatment by Barocas, Hardt, and Narayanan \cite{barocas_fairness_ml}
consolidates this measurement perspective. The shared lesson is that the system
under audit is the system as deployed, with its data, defaults, and interfaces,
not an abstraction of it. This thesis applies that lesson to LLM access
channels: the consumer app is the deployed condition most users encounter, and
it is audited as such rather than abstracted into ``the model.''

The broader critical literature supplies the motivation for treating deployed
products with suspicion of convenient assumptions. Gillespie
\cite{gillespie_algorithms} analyzes how algorithmic systems acquire public
relevance and authority; Burrell \cite{burrell_black_box} distinguishes forms of
opacity, including deliberate corporate secrecy, that block internal
inspection; Noble \cite{noble_algorithms} and Eubanks \cite{eubanks_automating}
document harms from deployed ranking and eligibility systems on real
populations; and Crawford \cite{crawford_atlas} situates AI systems in their
material and political supply chains. Closer to the interface, Amershi et al.\
\cite{human_ai_guidelines} codify how user-facing AI products shape interaction
through interface defaults, and Green and Chen \cite{green_limits} show
empirically that the same algorithmic output can lead to different outcomes
depending on how humans receive and use it. Both points anticipate a core claim
of this thesis: the surface through which an answer is delivered is part of the
behavior worth auditing, not packaging around it.

\subsection{From Documentation to Black-Box Inspection}

One response to opaque systems is standardized documentation. Model cards
\cite{model_cards} ask providers to report intended use, evaluation conditions,
and known limitations; datasheets for datasets \cite{datasheets} and data
statements \cite{data_statements} do the same for training data. These
artifacts are valuable, but they describe the model or dataset as an artifact,
not the productized deployment a user reaches: a model card rarely specifies
the hidden system prompt, tool wiring, retrieval behavior, or interface
rendering of a consumer app, and these are exactly the layers this thesis finds
to matter observably. Documentation and auditing are therefore complements: the
former states what a provider intends, the latter checks what a deployment
does.

A second response is post-hoc explainability. Local surrogate methods such as
LIME \cite{lime} and Shapley-value attribution \cite{shap} explain individual
predictions of accessible models. For closed commercial LLM deployments these
techniques have limited reach: there is no feature space or gradient access at
the product boundary, and the deployment stack can change without notice. The
practical alternative for proprietary conversational systems remains controlled
input--output probing, which is the method adopted here. The thesis accordingly
makes no interpretability claims; it codes observable outputs under documented
access conditions.

\subsection{API and WebUI Access-Channel Studies}

The closest direct precedent for this thesis is work on API/WebUI content
moderation disparities. Lipphardt et al.\ \cite{dualstandards} examine content
moderation differences between API and WebUI interfaces in large language models
and show that the access route itself can be an empirically relevant condition.
Their study matters because it shifts the evaluation question from only
``which model is being tested?'' to also ``which deployed access surface is
being tested?'' This is the same methodological move made in the present thesis.

The contribution here is narrower and complementary. Rather than focusing only
on content moderation disparity, this audit compares a fixed 60-prompt bank
across the OpenAI API and the official ChatGPT app, then codes refusal status,
safety framing, factuality, format compliance, unsupported inference, and
visible UI-surface signatures. This makes it possible to separate several
observable deployment-surface differences: bare refusal versus safe redirection, visible
sources, local context, and factual support. Dual Standards is therefore a
related-work anchor, not a prompt source or replacement for the paired local
audit.

\subsection{LLM Development and the Deployment Stack}

The systems audited here descend from a short but consequential line of work.
Brown et al.\ \cite{gpt3} showed that a sufficiently large autoregressive
language model can perform many tasks from prompts alone, which made the
prompt, rather than task-specific training, the interface to model behavior.
Ouyang et al.\ \cite{instructgpt} then aligned such models to human preferences
with reinforcement learning from human feedback, defining the
instruction-following assistants that consumer products now expose. Anthropic's
work on assistant alignment formulated the helpful--honest--harmless framing
\cite{helpful_harmless} and proposed Constitutional AI \cite{constitutional_ai},
which trains harmlessness against an explicit set of principles. LaMDA
\cite{lamda} is an early example of a dialog system whose safety and
groundedness were tuned as product properties, including retrieval-style
grounding. The significance for this thesis is that alignment and safety
behavior are trained, provider-specific properties layered onto the base model:
there is no single canonical ``model behavior'' that an access channel merely
transmits.

Documentation of recent frontier systems makes the deployment layering
explicit. The GPT-4 technical report \cite{gpt4_report} describes
post-training, safety mitigations, and system-level interventions added between
the base model and the served product. Meta's Llama 2 and Llama 3 reports
\cite{llama2,llama3} document safety fine-tuning and release accompanying
system-level safeguards; because their weights are open, the gap between a raw
model and a productized deployment is directly visible in those ecosystems. For
closed systems, the same layering exists but is observable only from the
outside, which is precisely the audit problem studied here.

This line of scaling also generated the critical literature on its risks.
Bender et al.\ \cite{stochastic_parrots} question what scaled language models
actually learn and warn against mistaking fluency for grounded competence;
Bommasani et al.\ \cite{foundation_models} name the homogenization risk of many
products inheriting the behavior, and the flaws, of a few foundation models;
and Weidinger et al.\ \cite{weidinger_risks} taxonomize the resulting harm
classes. The homogenization point sharpens the access-channel question: when
one provider's model family backs both an API and a consumer app, evaluators
are tempted to treat results from one surface as covering the other. The
present results show that this inference is not free.

\subsection{Benchmark Evaluation and Its Access Assumptions}

Modern LLM evaluation is organized around shared benchmarks. MMLU \cite{mmlu}
measures multitask knowledge, GSM8K \cite{gsm8k} grade-school mathematical
reasoning, and HumanEval \cite{codex} program synthesis; BIG-bench
\cite{bigbench} aggregates hundreds of heterogeneous tasks, and HELM
\cite{helm} deliberately standardizes scenarios, metrics, and conditions so that
models can be compared holistically. Instruction-tuned and chat-style systems
added their own evaluation layer: Self-Instruct \cite{self_instruct} and
Alpaca \cite{alpaca} demonstrate how cheaply instruction-following variants can
be produced from a base model, and MT-Bench \cite{mtbench} and G-Eval
\cite{geval} use strong LLMs as judges of open-ended responses.

Two properties of this ecosystem motivate the present audit. First, almost all
of these benchmarks are executed against programmatic endpoints under scripted
conditions, because that is what scales; the benchmark literature is, in
effect, an API-condition literature. HELM makes the standardization explicit
and is transparent about it, but the consumer surface remains outside the
standardized frame. Second, the proliferation of cheaply derived variants and
judge models means that ``the model'' named in a benchmark result is already a
bundle of base model, tuning data, and serving configuration. This thesis takes
the next step and treats the full deployment stack, including the consumer
interface, as part of what a behavioral measurement is about.

\subsection{LLM Safety and Refusal Evaluation}

Safety evaluations ask whether models refuse harmful instructions while still
answering benign requests. HarmBench \cite{harmbench} provides a standardized
framework for automated red-teaming and robust refusal evaluation. Do-Not-Answer
\cite{donotanswer} curates prompts that responsible models should decline and
supports low-cost automatic evaluation. MLCommons' AILuminate effort
\cite{ailuminate} formalizes hazard categories for product-level risk and
reliability assessment. These projects are important because they show how
unsafe or sensitive behavior can be turned into repeatable prompt sets.
Moderation classifiers such as Llama Guard \cite{llama_guard} additionally show
that providers can interpose a separate safety model between user and LLM; from
a black-box perspective, such components are part of the deployment stack and
indistinguishable from the model itself, which is one more reason to audit the
deployed bundle rather than assume a bare model.

Recent refusal and jailbreak-evaluation work reinforces the same design space.
SORRY-Bench \cite{sorrybench} systematically evaluates safety refusal,
StrongREJECT \cite{strongreject} provides a jailbreak/refusal evaluation
benchmark, and WildGuard \cite{wildguard} connects prompt harmfulness, response
harmfulness, and refusal detection in an open moderation tool. RefusalBench
\cite{refusalbench_bio} further illustrates why refusal rate alone can misstate
calibration: in biological-research prompts, strict refusal varies widely across
models and can penalize safe, useful engagement. These sources help position the
refusal labels and redaction policy, although the final prompt bank for this
thesis uses Do-Not-Answer and HarmBench for its high-risk refusal-expected rows
rather than importing raw prompts from all available refusal benchmarks.

For this thesis, these benchmarks provide two kinds of structure. First, they
identify prompt families where refusal is expected. Category B uses
refusal-expected prompts from Do-Not-Answer and HarmBench, with raw high-risk
text kept only in private local files. Second, they motivate boundary prompts
where the right answer is neither unrestricted compliance nor blanket refusal.
For example, educational cybersecurity, self-harm support, public-safety
escalation, and news-literacy prompts are answerable but require safe framing.
These prompts form Category C.

The deployment-surface question adds a further layer. A refusal benchmark run through
an API may measure one safety posture, while a consumer application may add
safe-completion language, visible sources, local emergency context, or different
refusal granularity. The key issue is therefore not only whether a model refuses,
but how refusal is expressed through a particular access surface.

\subsection{Red-Teaming and Jailbreak Research}

A complementary literature attacks safety behavior rather than measuring it.
Ganguli et al.\ \cite{ganguli_redteam} describe large-scale manual red-teaming
of assistant models and the lessons of scaling it; Perez et al.\
\cite{redteam_lms} automate the search for failure-inducing prompts using
language models themselves. On the adversarial side, Zou et al.\
\cite{universal_attacks} construct universal adversarial suffixes that transfer
across models, Chao et al.\ \cite{pair} show black-box jailbreaks can be found
in tens of queries, and Shen et al.\ \cite{do_anything_now} characterize the
in-the-wild ``DAN'' prompt ecosystem. Wei et al.\ \cite{jailbroken} explain
such failures as products of competing training objectives and mismatched
generalization between capability and safety training.

This thesis is not an attack study: Category B uses straightforward harmful
requests, not adversarial wrappers, and the audit asks how each surface
declines rather than whether it can be broken. The jailbreak literature is
nevertheless relevant for one structural reason: attacks are developed,
reported, and patched against specific deployed surfaces, and transfer between
surfaces is an explicit research variable in that literature. That is an
implicit acknowledgment that the API condition and the product condition are
not interchangeable, which is the assumption this thesis tests directly with
benign instruments.

\subsection{Over-Refusal and Calibration}

Safety systems can fail in the opposite direction by refusing benign requests.
XSTest \cite{xstest} targets this problem by testing prompts that contain words
or structures associated with unsafe content but are clearly safe in context.
Examples include ordinary programming, sports, food, or photography uses of
terms that can have violent meanings. OR-Bench \cite{orbench} extends the
over-refusal problem to a larger benchmark setting.

Over-refusal is important for deployment-surface auditing because refusal calibration
may differ by surface. A provider may choose one refusal style for API access and
another for the consumer app. The app may also include more explanatory framing
or safe alternatives, which changes the coding boundary between a full refusal
and a partial safe redirect. Category A in the prompt suite draws on XSTest and
OR-Bench-style allowed-sensitive prompts to test whether one channel refuses or
over-warns where the other answers normally.

The corrected results of this thesis also show why calibration claims require
careful coding. In an earlier pass, some socially sensitive rows appeared to show
API over-refusal. After transcript correction and a stricter treatment of
``cannot determine'' answers, those rows became negative controls rather than
positive over-refusal evidence. The stronger refusal difference is in Category B:
the API more often gives a bare full refusal, while the app gives a bounded safe
redirect.

\subsection{Instruction Following}

Instruction-following benchmarks test whether a model satisfies explicit,
verifiable constraints. IFEval \cite{ifeval} is useful because many of its
constraints can be checked without subjective semantic judgment: output must be
lowercase, avoid commas, contain a certain number of placeholders, or follow a
specified format. Such prompts are valuable in a deployment-surface study because
they separate product-layer effects from broad answer quality. If the API and
app both understand the prompt but only one satisfies a strict format
constraint, the difference is visible and concrete.

The consumer-app setting complicates format evaluation. App responses may include
rendered markdown, visual separators, code blocks, or extra conversational
phrasing. A screenshot transcript may also introduce artifacts that look like
format failures. This is why the thesis treats format compliance conservatively.
The final analysis includes only one format-compliance divergence after
prompt-echo cleaning and screenshot correction. Cases that initially looked like
format failures, such as S035 and S038, were removed from the final claim once
the screenshot evidence was checked.

\subsection{Factuality, Hallucination, and Retrieval}

Factuality benchmarks test whether models provide correct answers to factual
questions. SimpleQA \cite{simpleqa} targets short-form factuality, while
TruthfulQA \cite{truthfulqa} examines whether models reproduce common falsehoods
or misconceptions. Beyond benchmarks, a substantial literature measures and
detects hallucination: Maynez et al.\ \cite{maynez_factuality} distinguish
faithfulness from factuality in generation, Ji et al.\
\cite{hallucination_survey} survey hallucination across natural language
generation, FActScore \cite{factscore} decomposes long-form output into
atomically checkable facts, and SelfCheckGPT \cite{selfcheckgpt} detects likely
hallucination from the inconsistency of multiple sampled outputs. The
sampling-consistency idea in SelfCheckGPT is methodologically close to the
within-channel repetition check used in this thesis: both treat agreement
across independent samples from the same system as an internal noise floor
against which other differences can be judged.

A parallel line of work changes the factual support of an answer by adding
retrieval. Dense passage retrieval \cite{dpr} and retrieval-augmented
generation \cite{rag} ground answers in retrieved documents; WebGPT
\cite{webgpt} couples a model to a browser and trains it to cite sources;
GopherCite \cite{verified_quotes} optimizes answers supported by verified
quotes; and retrieval augmentation measurably reduces hallucination in dialog
\cite{blenderbot2}. In API-only evaluation, factual answers typically reflect
parametric memory under the selected settings. In a consumer app, the visible
product may add exactly the retrieval and citation machinery this literature
describes. The deployment surface can therefore change not only presentation
but the factual support available to the user.

This thesis uses neutral factual controls for two reasons. First, they serve as a
non-safety comparison category. A surface difference that appears only in high-risk
prompts might be interpreted narrowly as safety policy; a surface difference on
factual controls points to a broader access-layer difference. Second, factual
controls make UI-surface signatures meaningful. If the app shows visible sources
and the API provides a memory answer that is coded as a possible mismatch, the
result is a concrete example of how the app surface can change the user's
evidence environment.

\subsection{Bias and Socially Sensitive Inference}

Measuring social bias in language technology predates chat assistants. Word
embeddings were shown to encode occupational and gender stereotypes
\cite{debiasing_embeddings}, and Caliskan et al.\ \cite{caliskan_bias}
demonstrated that human-like implicit biases are recoverable from ordinary
text corpora. For generative models, Sheng et al.\ \cite{sheng_bias} measured
biased ``regard'' toward demographic groups in open-ended generation, and
RealToxicityPrompts \cite{realtoxicityprompts} showed that innocuous prompts
can elicit toxic continuations. Masked and pretrained models are probed by
CrowS-Pairs \cite{crows_pairs} and StereoSet \cite{stereoset}. Blodgett et al.\
\cite{blodgett_survey} critically survey this literature and argue that
``bias'' claims need explicit normative grounding in who is harmed and how.

The BBQ benchmark \cite{bbq} provides socially sensitive question-answering
items where a correct answer often requires avoiding unsupported inference.
Rather than selecting a stereotyped candidate, the model should recognize when
the provided information is insufficient. This makes BBQ-style prompts useful as
controls for unsupported social inference. In this thesis, Category E adapts
gender, nationality, and disability-status prompts into recommendation-style
questions that ask the model to decide based only on the given information.

The important coding decision is that an insufficiency answer is an answer, not
a refusal. A model that says the information does not determine who held a role
or who understood a discussion is performing the expected behavior. Treating that
as refusal would inflate over-refusal counts and distort the deployment-surface
claim. The final corrected analysis uses Category E as a negative control: none
of the ten rows is counted as a substantive API/app divergence. This coding
choice follows the normative-grounding point of Blodgett et al.: stereotype
avoidance is the desired behavior in these items, so it must not be scored as a
failure mode.

\subsection{Evaluation Validity and Access Channels}

The benchmark families above define what to ask and how to score important
behaviors. They do not, by themselves, solve the deployment-surface problem. A
benchmark result always has a collection context: model version, sampling
settings, system prompt, tool availability, account state, interface, and
timestamp. API studies often document these details more cleanly than app
studies, but the cost is that they may not represent the consumer-app experience.

This thesis therefore treats deployment surface as part of evaluation validity. The
research question is not whether API studies are wrong. APIs are indispensable
for reproducibility and scale. The question is whether API results transfer to a
consumer app without further evidence. The results show that this transfer cannot
be assumed: the app can add visible sources, local context, and different refusal
granularity. These differences are concrete enough to matter for safety,
compliance, factuality, and user-facing audit conclusions.

\subsection{Positioning of This Thesis}

Relative to the benchmark literature, the contribution is not a new benchmark
score. The prompt bank is small and deliberately diagnostic. Relative to
algorithm auditing, the contribution is a concrete operationalization of
deployment surface for LLM systems, in the audit tradition of Sandvig et al.\
and the deployed-product focus of Gender Shades. The thesis defines a coding
variable, \texttt{ui\_surface\_signature}, that captures visible app-specific
affordances without inferring hidden mechanisms. It also documents a practical
workflow for combining API logs with screenshot-backed app evidence, transcript
QA, manual review, and paired statistical summaries.

The thesis is therefore best read as a controlled deployment-surface audit. It does
not claim broad population estimates over all prompts or all model families. It
shows that, on a benchmark-grounded prompt bank, a bare API condition and
productized app deployment can produce different observable behavior under
documented conditions. That finding is
enough to motivate a broader methodological recommendation: LLM audits should
report the access surface they measure, and API-only claims should not be
generalized to consumer applications without validation.
