\section{Future Work and Research Agenda}

The thesis has reached the point where the central empirical question is
answerable from the collected evidence: under the documented conditions, the bare
API condition and the official productized consumer app did not always produce
the same observable behavior. The remaining research agenda is therefore not a search for a new
thesis claim. It is a plan for strengthening, stress-testing, and extending the
claim while keeping the same methodological discipline. This section translates
the limitations and discussion into concrete next steps.

The guiding objective is to make deployment-surface auditing more comparable
across time, providers, and user-facing products. The current study shows that
the deployment surface can matter, but it does so with one complete OpenAI API/app
comparison, one prompt-bank version, one app run per prompt, API-only vendor
baselines, and API-side repetition checks for OpenAI and Claude. A stronger
follow-up program would preserve the paired black-box design while adding
app-side repeated measurements, broader prompt coverage, independent coding, and
more precise documentation of app-side context. The
sections below prioritize work that is feasible with the existing repository and
that would most improve the evidential value of the thesis.

\subsection{Replication and Statistical Power}

The first extension is repetition. The present design uses 60 paired prompts,
which is sufficient to demonstrate concrete cases of deployment-surface
divergence but not
enough to estimate stable category-level rates with narrow uncertainty
intervals. A replication pass should keep the same prompt IDs, categories,
expected-behavior labels, and coding guide, but collect multiple independent
runs per prompt and channel. A practical target is three to five repetitions per
prompt, collected over separate sessions or days. That would turn the current
60-pair comparison into 180 to 300 paired observations while preserving the
same within-prompt structure.

The replication should be designed around observable stability rather than only
aggregate counts. For each prompt, the analysis can distinguish three cases:
stable no-divergence, stable divergence, and unstable divergence. Stable
no-divergence means both channels repeatedly produce the same final label.
Stable divergence means the same difference recurs across repetitions, for
example a repeated full-refusal versus safe-redirect split. Unstable divergence
means at least one channel changes label across repetitions. This distinction is
important because deployment-surface differences are not necessarily
deterministic. A
consumer app may route differently when search is available, when local context
is inferred, or when the model decides to use a tool. The audit should therefore
measure both channel differences and within-channel variability.

The statistical analysis should also be pre-registered before collection. The
follow-up should specify one primary endpoint, such as paired full-refusal
asymmetry, and a fixed family of exploratory moderator tests with a stated
multiple-comparison rule. For refusal labels, repeated measurements can be
analyzed using prompt-clustered summaries: first compute the per-prompt frequency
of API-only refusal and app-only refusal, then test whether the paired difference
remains directionally consistent. For the broader substantive-divergence outcome,
a mixed-effects logistic model or a simpler prompt-level bootstrap would be
appropriate. The main thesis does not need such models, but a follow-up paper
would benefit from separating prompt effects from deployment-surface effects.
The current repository already contains scripts for exact tests and Wilson
intervals; the next step is to add repeated-run aggregation while keeping the
simpler tables for auditability.

Replication also reduces the risk that the result is driven by a single
collection moment. LLM products change frequently, and the app surface can change
without an obvious public version number. A repeated design should record the
date, local time, account state, browser or app version, visible model selector,
search/tool indicators, and any visible source widgets for every run. These
metadata do not reveal hidden routing, but they make later interpretation much
less ambiguous.

\subsection{Larger and More Balanced Prompt Bank}

The second extension is prompt coverage. The present 60-prompt bank is balanced
for a compact thesis: it covers allowed-sensitive over-refusal, refusal-expected
guardrailing, boundary safety framing, instruction following, social
recommendation controls, and neutral factual controls. This is enough to show
that deployment-surface differences appear across several behavioral surfaces.
It is not large enough to claim a comprehensive taxonomy of such effects.

A larger bank should retain the six current categories but add narrower
subcategories. In allowed-sensitive prompts, the current homonym-style examples
could be split into software/process, sports, legal/business, cooking, and
everyday language. In refusal-expected prompts, the private high-risk set should
be balanced across privacy, cyber, misinformation, illegal activity, harassment,
and crisis-support boundaries, while still keeping executable prompt text out of
shared artifacts. In boundary safety prompts, the next version should distinguish
health, legal, safety reporting, scam prevention, civic misinformation, and
fictional writing. These subcategories would make it possible to ask whether the
app/API difference is a general refusal-style difference or concentrated in a
few risk domains.

Instruction-following prompts also deserve a larger role. The current corrected
analysis found only one final substantive format case after screenshot QA. That
is a useful negative finding, but it rests on a small set of eight instruction
prompts. A follow-up should include objective constraints that are easier to
score automatically: exact token counts, JSON validity, required separators,
forbidden punctuation, case constraints, ordered lists, and multilingual output.
These prompts are valuable because they are less dependent on subjective safety
interpretation. If deployment surfaces diverge on such tasks, the evidence is
especially easy to audit.

The social recommendation category should be expanded carefully, not because the
current result is weak, but because it corrected an earlier mistaken direction.
After transcript QA, the social rows function as negative controls: both
channels generally avoided unsupported inference. A larger follow-up should
preserve that role and include more protected-attribute settings, but the coding
must remain conservative. The expected behavior should be refusal of unsupported
inference or explicit insufficiency, not moralizing language. This distinction
prevents the analysis from counting ordinary stereotype avoidance as
over-refusal.

\subsection{Consumer-App Baselines Beyond OpenAI}

The current vendor baselines are API-only. They are useful because they show that
raw provider APIs already differ in refusal posture on the same prompts, but they
do not answer whether Claude, Gemini, or other consumer apps exhibit their own
app/API differences. A direct extension would add official consumer-app
collection for those systems using the same prompt bank, the same screenshot
evidence protocol, and the same coding guide.

This extension should be explicit about comparability. Different providers expose
different consumer surfaces: model pickers, search toggles, workspace memory,
file handling, citation displays, safety notices, and regional defaults vary
across products. The goal should not be to force all consumer apps into a
fictional identical condition. Instead, the audit should document each access
surface and compare the effect of using the official consumer route versus the
programmatic route within each provider. The primary unit would become a
provider-specific API/app pair. Cross-provider claims would then be framed as
patterns across pairs, not as a single leaderboard.

Adding consumer-app baselines would also improve the interpretation of UI-surface
signatures. In the current OpenAI comparison, visible sources and local-context
signals are app-side signatures absent from plain API text. Other products may
surface different signatures, such as search summaries, suggested follow-up
cards, document canvases, or location widgets. The coding scheme should allow
provider-specific UI details while keeping a common top-level taxonomy:
source-visible, context-visible, tool-visible, file/canvas-visible, and no
visible app affordance.

\subsection{Longitudinal Drift Audits}

The third major extension is time. A one-time audit can identify channel
differences at the moment of collection, but it cannot tell whether those
differences are stable product properties or temporary artifacts of a release.
Longitudinal auditing is therefore necessary for any stronger claim about
consumer-facing reliability.

A feasible longitudinal design would rerun the same prompt bank weekly or
monthly for a fixed period. Each wave should freeze the prompt text, account
conditions, browser/app conditions, and API parameters as much as possible. The
analysis should report both wave-level divergence rates and prompt-level change
events. For example, a prompt could move from API-only refusal in one wave to
both-channel safe redirection in another. That change would matter even if the
aggregate rate stayed similar, because it would show that product updates can
shift the user-facing safety boundary.

Longitudinal work should also record version uncertainty honestly. APIs often
provide a visible model name, but consumer apps may expose only a product label.
Even when a label is visible, it may not fully specify routing, tool use, memory,
or moderation layers. The correct methodological response is not to infer hidden
versions, but to timestamp the collection and preserve screenshots. That makes
the audit reproducible as an observation of the access condition, even if the
internal mechanism remains unknown.

\subsection{Reliability and Coding Governance}

Independent reliability remains the cleanest way to quantify coding agreement,
but the current thesis no longer depends on it as a single point of failure. The
coding-dependence sensitivity analysis shows that no final substantive case is
safety-framing-only. Completing the independent coding route would still
strengthen the taxonomy because some final labels are interpretive: safe
redirection versus full refusal, light versus moderate safety framing, and
substantive versus minor divergence.

The reliability task should focus on labels that affect claims. It is less
important to recode every auxiliary note than to independently code refusal
status, safety-framing strength, format compliance, unsupported inference,
factuality, UI-surface signature, and substantive-divergence status. The coder
should receive the coding guide, prompt context, and raw outputs, but not the
primary labels. After coding, disagreements should be summarized by label type. A
low agreement score on a label does not automatically invalidate the study; it
identifies which claims need narrower wording or adjudication.

For future projects, the coding workflow should include a short calibration
round before the full audit. The coder pair can label ten prompts, discuss
disagreements, revise the guide, and only then code the final sample. That
process would improve reliability without retroactively changing the
pre-registered prompt bank. It would also make the distinction between minor and
substantive divergence clearer, which is the most judgment-sensitive part of the
current study.

\subsection{Mechanism-Probing Without Overclaiming}

The current thesis intentionally avoids causal claims about hidden prompts,
moderation layers, or routing. Future work can probe mechanisms more directly,
but only with designs that support such claims. One safe direction is controlled
ablation at the deployment surface. For the app, researchers can repeat prompts with
search disabled where possible, memory disabled where possible, location
permissions controlled, and fresh sessions. For the API, researchers can vary
system prompts, temperature, tool availability, and response-format settings.
These ablations would not reveal proprietary internals, but they would identify
which visible conditions are sufficient to reproduce certain differences.

Another direction is response-forensics coding. Instead of claiming a hidden
cause, the audit can code observable traces: citations, source snippets,
localized entities, tool-call notices, refusal templates, and answer structure.
If a factual answer is source-visible in the app and unsupported in the API, that
is a trace-level difference. If an app answer mentions a local city when the
prompt did not provide one, that is a context trace. Such coding keeps the
analysis grounded in evidence while still moving closer to explanation.

\subsection{User-Centered Consequence Measures}

The present study is output-centered: it compares response labels. A later study
could ask whether those differences matter for users. This is especially
relevant for boundary safety and factual controls. A safe redirect may be more
useful than a bare refusal if it gives legitimate next steps. A source-visible
answer may be more trustworthy if users inspect the sources, but it may also
create unwarranted confidence if sources are weak. The next layer of evaluation
should therefore include utility, perceived helpfulness, perceived safety, and
decision quality.

Such a study should be designed cautiously. It should not expose participants to
private high-risk prompt text or ask them to execute unsafe behavior. Instead,
participants could compare redacted or harmless outputs, judge whether an answer
is actionable in a legitimate way, or identify whether a response gives enough
information to make a safe decision. The user-centered layer would complement,
not replace, the black-box audit. The audit establishes that outputs differ; the
user study estimates whether those differences are consequential.

\subsection{Practical Priority Order}

The table below converts the research agenda into a priority order. The first
two items are most relevant for a stronger final thesis package. The remaining
items are better treated as follow-up research.

\begin{longtable}{L{0.06\linewidth}L{0.22\linewidth}L{0.27\linewidth}L{0.23\linewidth}}
\caption{Prioritized next steps for strengthening the deployment-surface audit.}\\
\toprule
Rank & Task & Immediate output & Why it matters \\
\midrule
\endfirsthead
\caption[]{Prioritized next steps (continued).}\\
\toprule
Rank & Task & Immediate output & Why it matters \\
\midrule
\endhead
\bottomrule
\endfoot
1 & Add app-side repetitions for key rows & Stability table for the 17 substantive cases and matched controls & Quantifies consumer-app stochasticity, the remaining repetition gap. \\
2 & Complete independent coding & Filled blind sheet and reliability report & Strengthens interpretive labels and supervisor-facing credibility. \\
3 & Add consumer-app baselines for other providers & Provider-specific API/app matrices & Tests whether deployment-surface differences generalize beyond one product. \\
4 & Expand instruction-following prompts & Larger objective-format subset & Tests a less subjective behavior class. \\
5 & Run longitudinal waves & Drift report by prompt and category & Measures whether observed effects persist over time. \\
6 & Add access-surface ablations & Controlled search, memory, location, and tool-condition notes & Narrows plausible explanations without claiming hidden mechanisms. \\
7 & Conduct user-centered evaluation & Helpfulness and safety-utility ratings for harmless output pairs & Connects observable divergences to user consequences. \\
\end{longtable}

\subsection{Highest-Value Immediate Extension: An App-Side Stability Panel}

If only one additional empirical sprint is feasible, the most efficient choice
is not a full new benchmark. It is a targeted app-side stability check around
the existing final cases. The API side already has
additional repetitions for the full 60-prompt bank; the remaining gap is the
consumer-app side. The sprint would recollect the 17 substantive divergence
prompts plus a matched set of non-divergent controls from the same categories in
the app under documented conditions. This would create a compact stability panel
without changing the research question or requiring a new prompt bank.

The output of that sprint should be a small row-level stability table. Each
prompt can be labeled as stable, partially stable, or unstable. Stable means the
same final divergence label appears in all repetitions. Partially stable means
the same direction appears more than once but at least one run changes label.
Unstable means the divergence disappears or reverses across repetitions. This
table would directly answer the most natural reader question after
the Results chapter: are the 17 cases isolated examples, or do they
repeat when recollected?

The quote-level screenshot audit already provides a
quality-control record for the current 17 cases. The same discipline should be
kept for any new app repetitions: each screenshot, transcript, and final coding
row should be checked together before adding the result to a table.

The same sprint should complete the independent-coding route when a second
coder is available. The blind coding sheet already exists, so the bottleneck is
not infrastructure but execution. If the coder can complete only a subset, the
subset should prioritize the 17 substantive cases and matched controls. A
smaller honest reliability report is more valuable than a broad but rushed one.
Absent a second coder, the documented sensitivity analysis and screenshot-QA
route remain the stated evidence basis, without implying stronger agreement
evidence than exists.

These three tasks form a realistic bridge between the present thesis and a more
publishable audit: stability checks show whether cases repeat, screenshot QA
protects the evidence chain, and reliability work tests the coding scheme. None
of them requires weakening the current claim or adding unsafe prompt text to the
public thesis artifact.

This agenda keeps the thesis objective intact. The central claim is already
supported at the level appropriate for a black-box master's thesis: a bare API
condition is not automatically a proxy for a productized consumer-app deployment.
The next work should make that claim more stable, more general, and more useful
for audit practice.
