\documentclass{ddis-thesis}

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{ifthen}
\usepackage{pdflscape}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{float}
\usepackage{placeins}
\usepackage{xltabular}
\usepackage{makecell}
\usepackage{adjustbox}

\setlength{\emergencystretch}{3em}
\setlength{\tabcolsep}{3pt}
\newcolumntype{L}[1]{>{\raggedright\arraybackslash}p{#1}}

\newcommand{\studentName}{Rakhmatillokhon Khoshimov}
\newcommand{\studentEmail}{rakhmatillokhon.khoshimov@uzh.ch}
\newcommand{\studentNo}{23060361}
\newcommand{\studentProgram}{MSc Artificial Intelligence}
\newcommand{\thesisTitle}{Auditing Access Channel Effects in LLM Behavior}
\newcommand{\thesisType}{Master's Thesis}
\newcommand{\submissionDate}{1 July 2026}
\newcommand{\advisorName}{Dr. Aleksandra Urman}
\newcommand{\responsibleProfessor}{Prof. Dr. Anik\'o Hannak}
\newcommand{\departmentName}{Department of Informatics of the University of Zurich}
\newcommand{\scgUrl}{www.ifi.uzh.ch/scg}

\author{\studentName}
\title{\thesisTitle}

\begin{document}
\pagenumbering{roman}

\hypersetup{pageanchor=false}
\ddistitlepage
\cleardoublepage
\setcounter{page}{1}
\hypersetup{pageanchor=true}

\begin{abstracten}
Large language model audits are often conducted through provider APIs because
API access is reproducible and easy to automate. Ordinary users, however,
interact with nominally similar model families through consumer applications
that may add system instructions, tools, retrieval, session handling, location
context, and visible interface affordances. This thesis asks whether conclusions
from a bare API condition transfer to a productized consumer-app deployment. The
design does not isolate access channel as a causal mechanism: it compares the
OpenAI Responses API (\texttt{gpt-4o}, no custom system prompt) with the official
ChatGPT app, whose archived screenshots do not reliably preserve an in-frame
model label and whose hidden system prompt, tools, retrieval, routing, and
session state cannot be separated from the app condition. Using a frozen
six-category, 60-prompt suite grounded in recognized benchmarks, each response is
coded for refusal status, safety framing,
instruction following, unsupported inference, factuality, verbosity, and visible
UI-surface signatures. Manual review identifies 17 substantive observable
divergences. Ten of these have response-core anchors: six refusal-boundary
cases, three factual-support cases, and one format-compliance case. The
remaining seven are product-surface or context cases, such as visible sources or
localized context, and are treated as methodologically relevant but less
theoretically surprising. The primary refusal asymmetry is directional but not
conventionally significant (McNemar exact $p=0.062$, one discordant pair below
the exact-test floor), and all moderator p-values are exploratory pattern checks
on small cells rather than confirmatory evidence. Two additional API repetitions
show that the OpenAI API refusal label is perfectly stable within that channel
(Fleiss $\kappa=1.0$), and on every one of the 60 prompts the API's independent
re-runs resemble each other more than the API resembles the app (mean token
Jaccard 0.47 versus 0.22), so API-side sampling noise does not account for the
observed API--app differences. A targeted ChatGPT-app repetition check on the
17 final cases adds two fresh app runs per prompt: 14 of the 17 original core
anchors reproduce in both repetitions, one reproduces in one of two, and two do
not reproduce. Crucially, all six refusal-boundary cases and all three
factual-support cases reproduce in both app repetitions, while the less stable
cases are the single format-compliance case and two visible-source-only
affordance cases.
An independent blind second-coder pass on 54 coding units shows strong agreement
for refusal status, format compliance, unsupported inference, and acceptable
agreement for factuality; UI-surface agreement is moderate, while
safety-framing strength is weak (Cohen's $\kappa=0.06$, weighted
$\kappa=0.27$). The thesis therefore treats safety framing only as descriptive
context. A coding-dependence sensitivity check further shows that no final
substantive case rests on safety framing alone, and every final case has a
non-framing evidence anchor such as visible UI surface, factual-control support,
exact format compliance, or a refusal-boundary contrast. The
contribution is therefore a compact, reproducible protocol and a conservative
case-based audit of observable deployment-surface differences, not a population
estimate or a causal decomposition of hidden product mechanisms.
\end{abstracten}

\begin{acknowledgements}
I thank Dr. Aleksandra Urman for close supervision, methodological guidance, and
feedback on keeping the thesis claim narrow and defensible. I also thank Prof.
Dr. Anik\'o Hannak for serving as the responsible professor and for supporting a
Social Computing Group project on auditing access-channel effects. The empirical
work in this thesis benefited from discussions about benchmark selection,
consumer-app evidence, coding discipline, and the risks of over-interpreting
black-box outputs. Any remaining mistakes in the analysis, coding, or prose are
my responsibility.
\end{acknowledgements}

\tableofcontents
\clearpage
\listoftables
\clearpage
\listoffigures
\clearpage

\pagenumbering{arabic}
\glsresetall

\input{sections/01_introduction}
\input{sections/02_related_work}
\input{sections/03_methods}
\input{sections/04_results}
\input{sections/05_discussion}
\input{sections/06_limitations}
\input{sections/07_future_work}
\input{sections/07_conclusion}

\phantomsection
\addcontentsline{toc}{chapter}{\bibname}
\printbibliography

\cleardoublepage
\phantomsection
\addcontentsline{toc}{chapter}{List of Acronyms}
\chapter*{List of Acronyms}
\glsaddall
\printglossary[type=\acronymtype]

\appendix
\input{sections/08_appendix}

\cleardoublepage
\hypersetup{pageanchor=false}
\pagestyle{empty}
\declaration

\end{document}
